---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---



Check if all packages required for this script has already been installed in the local computer. Otherwise, install any required packages.

```{r installpackages,results="hide"}
    if(!require("lubridate"))
        install.packages("lubridate")
    
    if(!require("dplyr"))
        install.packages("dplyr")


    if(!require("ggplot2"))
        install.packages("ggplot2")

    if(!require("knitr"))
        install.packages("knitr")

    
```



## Loading and preprocessing the data


Set global variables for echo = True

```{r setoptions, echo=TRUE, results="hide"}
    opts_chunk$set(echo=TRUE)
```

1. Code for reading in the dataset and/or processing the data

```{r readingData, results='hide'}
    
    activity <-  read.csv("activity.csv", stringsAsFactors = FALSE)
    activity$date <- as.Date(activity$date)
    activity <- tbl_df(activity)
```

## What is mean total number of steps taken per day?

2.  Histogram of the total number of steps taken each day         


```{r Qtn2}
  act_date <- activity%>%
              group_by(date)%>%
              summarise(no_steps = sum(steps,na.rm=TRUE), 
                        mean_steps = mean(steps,na.rm=TRUE))
    
  hist(act_date$no_steps,
       main = paste("Total Number of Steps Taken per Day"), 
       xlab="Number of Steps in any particular day")
```

3.  Mean and median number of steps taken each day
    
```{r Qtn3}
  print(paste("Mean number of steps taken per day: ", mean(act_date$no_steps,na.rm = TRUE)), digits = 2)
  print(paste("Median number of steps taken per day: ", median(act_date$no_steps,na.rm = TRUE)),digits = 2)
```





## What is the average daily activity pattern?
4. Time series plot of the average number of steps taken

```{r Qtn4}
  
    plot(act_date$date,act_date$mean_steps,type = "l", main = "Average Daily Activity Pattern",
       xlab = "5-Minute Interval", ylab = "Average steps taken")
```

5. The 5-minute interval that, on average, contains the maximum number of steps

```{r Qtn5}
  
    
  day_highest_mean <- as.matrix(act_date[act_date$mean_steps== max(act_date$mean_steps, na.rm = TRUE) & !is.na(act_date$mean_steps),"date"])
  
  print(paste("Date with highest average number of steps: ",day_highest_mean))
```



## Imputing missing values

6. Code to describe and show a strategy for inputing missing data

```{r missingData}  
  no_NA <- sum(is.na(activity$steps))
  print(paste("Total number of missing values in the dataset:",no_NA))
  
  # if steps is NA, then assign the mean of the date
  
  activity_new <- merge(activity,act_date, by="date")
    
  for(i in 1:nrow(activity_new)){
      if(is.na(activity_new$steps[i])){
          if(is.na(activity_new$mean_steps[i])){
                activity_new$steps[i] <- 0
          } else{
                activity_new$steps[i] <- activity_new$mean_steps[i]
          }
      }
  }
  
  activity_new1 <- activity_new %>%
                  select(date, steps,no_steps, interval, mean_steps)%>%
                  group_by(date)%>%
                  summarise(no_steps = min(no_steps), no_steps_new = sum(steps) ,mean_steps = min(mean_steps), new_mean_steps = mean(steps), difference = new_mean_steps-mean_steps)
```


7. Histogram of the total number of steps taken each day after missing values are inputed                                     
```{r Qtn 7}
    hist(activity_new1$no_steps_new,
       main = paste("Total Number of Steps Taken per Day"), 
       xlab="Number of Steps in any particular day")
```





## Are there differences in activity patterns between weekdays and weekends?

8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r Qtn8}
    activity_wdays <- activity_new
  
  activity_wdays$wday <- "weekday"
  
  for(i in 1:nrow(activity_wdays)){
        if(weekdays(activity_wdays$date[i],abbreviate = TRUE) %in% c("Sat","Sun")){
          activity_wdays$wday[i] <- "weekend"          
        }
  }
  
  
  activity_wdays <- aggregate(steps ~ interval + wday, data=activity_wdays, mean)
  
  ggplot(activity_wdays, aes(interval, steps)) + geom_line() + facet_grid(wday ~ .) +
    xlab("5-minute interval") + ylab("Average number of steps")

```  
